Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-22 src/article.tex
Updates from Overleaf

Technological advances have transformed how museums document, present and interpret their collections, fostering immersive experiences through tools such as laser scanning, 3D printing, and virtual reality \cite{allard2005use,Wachowiak01082009,RCM_2024_3D,Kuzminsky_LaserScan_2012,Schaich_3D_2007}. These technologies place an emphasis on experiential authenticity, enabling encounters that evoke the sensory, emotional, and intellectual essence of the past \cite{trant_Auth_1999}. However, as Pine and Gilmore note \cite{pinegilmore_2007}, achieving authenticity requires museums to navigate the delicate balance between preservation and meaningful engagement—a challenge that is particularly evident in the case of historical musical instrument collections \cite{McAlpine2014}.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{src/images/details.jpg}
\caption{Details of the replica, including the key slots and sides, keys, and jacks with plectra.}\label{fig:details}
\end{figure*}

@@

The final design is a 49-key harpsichord replica equipped with an optical sensor system to reproduce the tactile characteristics of the original instrument, visible in Figure \ref{fig:teaser}. Several key features were implemented to address the project’s requirements:

@@

A modular system of PCBs was designed to manage the sensors and process their output. The 49 QRE1113 sensors were distributed across seven boards, containing 7 sensors each. The PCBs were secured to the underside of the keyboard mechanism, allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

The project built upon earlier NIME research on generating MIDI messages from piano keystrokes \cite{McPherson2013}, adapting it to address the unique characteristics of harpsichords and the specified constraints. Whereas the previous design emphasised continuous gesture tracking, this implementation required discrete key-triggered data to align with the needs of MIDI-triggered audio playback. The adoption of gradient stickers in place of continuous optical tracking embodies this adaptation, striking a balance between precision, cost-efficiency, and practicality.

@@

The keyboard was designed to replicate the tactile and aesthetic sensations of playing an antique Italian harpsichord. Traditional materials were used, including walnut for the wrestplank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the Alessandro Trasuntino instrument at the San Colombano Museum. Details are visible in Figures \ref{fig:teaser} and \ref{fig:details}.

The rectangular poplar frame deviates from the traditional logarithmic form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string orders, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations.

The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile and auditory qualities.

\section{Hardware Design}

For the first iteration the data from the sensor would simply be used to test when a threshold had been passed. On passing the threshold a message would be sent to trigger the playback of a corresponding note.

For the initial prototyping stage, modifying an existing harpsichord was considered though ultimately discarded. An existing instrument would
have provided a test for scaling the electronics, but internal measurements and layout would have been too different.

Using a similar approach to Timmermans' Haptic Key project, \cite{Timmermans2020} our first step was to fabricate or procure a model of a single working key. San Colombano was able to provide a model 3-key of a harpischord mechanism by Graziano Bandini (Figure \ref{fig:3key}).

@@

A set of criteria was drawn to which the final sensor system would have
to satisfy. This criteria was used as a means of evaluating the
feasibility throughout the prototyping phase.

\begin{itemize}
\item
  Non-invasive: The sensor system should not require major modifications
  to the interface on which it is installed.
\item
  Low Latency: Reading sensors and then triggering an output should have
  a low latency. Following \cite{Jack2016} an latency was set as 10 ms
  with an upper limit set to 25 ms beyond which the sensors would be
  deemed unusable
\item
  Reliable: Data obtained should be reliable and accurate enough not to
  result in false positives or false negatives without extensive
  filtering
\item
  Scalable: The system should be scalable both economically and
  temporally. Given the open source nature of the project, it could not
  be prohibitively expensive to carry out. Also any system that would
  work on a single key would have to scale financially to potentially 50
  or more. The time required to install and calibrate the sensors would
  also have to scale.
\item
  Expandable: Any sensor system should be functionally expandable to
  allow for the usage of the data for control over other parameters.
\end{itemize}

The sensor system was designed so that they would not require any equipment outside what can commonly be found in a maker space at a university for assembly.

This decisions was motivated by the project's goal to be open source and reasonably recreated without procurement of any additional specialist equipment.

The electronics system can be divided into two designs, the "sensor board" (Figure \ref{fig:7-sensor-board}) for components for detecting jack displacement and controller board for collecting and processing sensor signals.

\begin{figure}  
  \centering
  \includegraphics[angle=-90,width=\linewidth]{src/images/7-sensor-board-v0.1.0.png} 
  \caption{} 
  \Description{} 
  \label{fig:7-sensor-board}
\end{figure}

\subsection{Sensor board}\label{sensor-board}

The 49 optical sensors are divided across 7 PCBs each containing

\begin{itemize}
\item
  x7 QRE1113 optical sensors
\item
  x7 100 $\Omega$ resistors
\item
  x7 10 k$\Omega$ resistors \footnote{a update on the design showed these could be reduced to a single resistor.}
\item
  x1 CD4051BE Multipexer
\end{itemize}

A number of sensors were tried before the final system was decided upon. 
Sensors tested for the design included, a 6-degrees-of-freedom inertial measurement unit, reed and hall sensors paired with magnets, and light-dependant resistors mounted to the jack rail.
None of the above satisfied the sensor criteria described and were thus discounted.

Following from the work in \cite{McPherson2013, McPherson2019}, we tested
a system using the Fairchild QRE1113 \footnote{Fairchild, now ON Semiconductor, QRE1113 datasheet \url{https://www.onsemi.com/download/data-sheet/pdf/qre1113-d.pdf}}. The
QRE1113 is a combination infrared LED and phototransistor sensitive to
IR light. The phototransistor of QRE1113 provides data on how much light
is present and in particular how much light is being reflected from a
surface. Since reflected light will be proportional to the distance of
a surface it is a good, close-proximity, distance sensor. Distance is
the way in which the instruments in \cite{McPherson2013, McPherson2019}
function as well the Moog Piano Bar from which it took inspiration.

The limitation for taking a distance approach is that there is not one
but 2 jacks per key. For each jack to be measured independently the data
would need to come from the jacks themselves.

% They \emph{could} be used and individual trigger points set in
% calibration, but this creates new problems.

\begin{figure}  
  \centering
  \includegraphics[,width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg} 
  \caption{} 
  \Description{} 
  \label{fig:cross-section}
\end{figure}

The optical sensors were directed at a greyscale gradient printed on a sticker and applied to the side of the jack.
This was first trialled using a generic inkjet printer and double-sided tape. 
For the final iteration, the stickers were printed on Coala 1D 100 Gloss PG 100 $\mu$m white gloss monomeric PVC with permanent grey acrylic dispersion adhesive \footnote{Datasheet: \url{https://www.antalis.co.uk/mediashare/g4media/pdf/TS_EN_COALA_1D_100_GLOSS_P_00_ISS_15022023.pdf}} using a HP Latex 115 and then cut with a Summa 150 roll cutter. 
The printer and roll cutter greatly reduced the manual labour for cutting stickers meaning the process would scale from the 6 jacks of the 3-key to the 98 jacks of the 49-key model.

\begin{figure}  
  \centering
  \includegraphics[angle=90,width=\linewidth]{src/images/jack-w-tag.jpeg} 
  \caption{} 
  \Description{} 
  \label{fig:jack-w-tag}
\end{figure}

During testing it was found that their was a considerable amount cross-talk between adjacent sensors. 
This meant the reading from key would be affected by the movement of the key next to it. 
The cross-talk was assumed to be as a result of light reflecting from other jacks.
Given that IR is used this was difficult to confirm by sight. 
A set of baffles was designed for the PCB that would wrap round the jack and stop light from other jacks or sensors \ref{fig:baffles}. 
The baffles were printed on a BAMBU X1 3D printers using PLA plastic. A dark pigmented PLA was used to avoid IR light simply being diffused. 
After the baffles were fitted, the cross-talk was eliminated.

\begin{figure}  
  \centering
  \includegraphics[,width=\linewidth]{src/images/baffles.png} 
  \caption{} 
  \Description{} 
  \label{fig:baffles}
\end{figure}

Calibration to pluck thresholds were to be carried out manually.  To allow for easy identification of the sensor being addressed during the calibration process, RGB LEDs were integrated into the design. The LEDs were also used in the identification of any malfunctioning or un-calibrated sensor.

TLC5940 or other PWM driver such as in \cite{McPherson2013} add robustness should an LED fail, but complexity to assembly process. 
Instead, the design uses addressable LEDs with integrated driver that share the same data line. 
This has the limitation where a faulty LED will stop all subsequent LEDs in the line from working.

The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
Transistor / Op-amp circuit could have been used similar to the one in \cite{McPherson2013}.

@@

Controller board designed around Arduino nano form factor. The Nano is
connected through headers to allow easy switching between chips. Should
another chip in a Nano for factor be found it could supplant the current
Nano without any need for a redesign.

The first version of the board exclusively used 2.54 mm pitch headers.
The reasons for this were flexibility when creating cable looms and to
leave open the possibility of rewiring the PCB.

This was the case for the LEDs as they initially were to be powered on
5V. It was discovered that the 3V regulator of the nano was sufficient
to power all LEDs at a lower brightness while reducing current draw.
Since brightness was not important.

The 2.54 mm pitch headers had two major drawbacks. First, the orientation
of the cables could be reversed, which in the best case scenario would
simply mean the system did not work correctly and in the worst case it
could potentially damage the components.

Secondly, the headers do not connect securely and wait of the cabling
was enough for components to become disconnected as the harpsichord was
handled.

To address this, once the wiring for the controller board was finalised
a new design was drafted this time using JST-PH connectors. No
commercial cables were immediately available in the configuration and
length required, as such a cable loom needed to be made which meant
crimping cables by hand. A crimping tool from JST is prohibitively
expensive compared to the cost of other components in the project
\footnote{from RS \url{https://uk.rs-online.com/web/p/crimp-tools/6880877}
£508 (inc. VAT)}. The crimping tool also appeared to be uncommon and was
not readily available in the workshops at the project's disposal.
Eventually a tool was found, but an alternative approach was also
researched in order to maintain the criteria of accessibility.

A locator was adapted from a creative commons STL model \footnote{\url{https://www.thingiverse.com/thing:1646016}} that allowed a generic crimping tool to be adapted for use with JST-PH crimps.

\subsubsection{Arduino Nano BLE}\label{arduino-nano-ble}

Arduino Nano BLE was chosen for a few reasons. The Nano form factor was
appealing as it would allow for easy changes between chipsets without
requiring any rewiring and thus no new PCBs would need to be created if
a better alternative was found later in the project. Another benefit to
the Arduino Nano boards was that typically had native-USB meaning the
could be programmed as a hardware USB MIDI device directly.

The Nano BLE was used in the initial stages for its integrated inertial measurment unit.

When it was finalised that the project would proceed using the QRE1113 a
number of microcontroller units (MCU) were considered.

The MCUs considered were the Arduino Nano 33 IoT, Arduino Nano 33 BLE,
Arduino Nano ESP32 and STM32 NUCLEO-L031K6.

A rough performance test was later carried out using an STM32
NUCLEO-L031K6, a Nano ESP32 and the BLE. The Nano BLE was found to be
more performative. To execute a single sensor read cycle of the firmware
the STM32 took 40 ms, and both the Nano ESP32 and Nano BLE required 11 ms
on average. The Nano ESP32 had high variability so it was decided to
continue with the Nano BLE.

BLE functionality also meant if the cable connection between the nano
and computer running audio synthesis was impractical, BLE MIDI was
available as a fallback. The Nano BLE also has a 12-bit DAC, which
provided a contingency should the default 10-bit not be sensitive enough
for the signal from the sensors.

Initially a Nano IoT was used for both WiFi and Bluetooth Low Energy
functionality, but it was found that 2 ADC channels were in fact not
useable and would have required additional multiplexers.

\subsubsection{EEPROM / NVRAM / FRAM}\label{eeprom-nvram-fram}

There would need to be the ability to store values on non-volatile
memory so that the electronics could be powered-down without losing
data.

Adjusting the values through firmware would not have been flexible or
practical. Instead an Ferroelectric RAM unit was chosen and connected
using an SPI interface. NVRAM or EEPROM would have sufficed and the FRAM
was primarily chosen for it's ready availability.

Most microcontrollers have some variety of non-volatile program memory
that can be read and writ accessed but in most cases this memory is
cleared during compilation and re-flashing. The small cost was far
outweighed by the potential for data loss if using onboard RAM.

SD card would provide an easier way to interact with calibration data.
The position of the controller board did not permit easy access.
Reliability and speed was preferred over storage capacity.

\subsection{Power}\label{power}

All electronics combined draw around 1.1 A of current at 5V with some fluctuation during start-up.

Arduino BLE has some functionality disabled to limit current draw.

Ideally the unit could be bus powered from USB.

The sensors are in a state of constantly drawing power. 
It is possible that powering only when data is required could be reduce power enough.

Current designs have a separate MIDI device for each jack row. 
Combining the devices and also reducing current draw to below the common 0.5 amps for a USB port would certainly be a challenge.

\section{Firmware Design}\label{firmware-design}

Arduino platform chosen for wide adoption, relative ease of use and
library support for components.

Firmware checks components are connected then reads the FRAM to see if
threshold data is already present. The rotary encoder and it's momentary
switch are polled for any change in state. A key is selected with the
rotary encoder while the momentary switch is used for changing between
states for key selection and threshold adjustment A double click of the
rotary encoder writes the current threshold values to FRAM. The sensors
for the current multiplexer channels are polled. If any sensor reads
above the threshold when it was previously below a MIDI On message is
sent. For the reverse case where the value was below the threshold when
it was previously above, a MIDI Note Off message is sent.

\section{Implementation}\label{implementation}

\begin{figure}  
  \centering
  \includegraphics[,width=\linewidth]{src/images/exhibition-user-1.jpeg} 
  \caption{} 
  \Description{} 
  \label{fig:exhibition}
\end{figure}

\subsection{Installation}\label{installation}

The electronic components were partly pre-assembled at the university of
Edinburgh. Though all measurements were known ahead, enough flexibility
was needed to allow for problems to solved in situ.

The final assembly took place at the NEMUS lab at the University of
Bologna over the final week in October 2024 for delivery at San
Colombano for exhibition in the first week of November in 2024.

A misreading in the original clearance between the keys and the PCB
meant that the keys were blocked entirely from moving. A 10 mm by 15 mm
section was milled out of each key to create extra
clearance. The milling required the removal, shortening and refitting of
the leather pads for the jacks. The extra work
added an extra day to the full install.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/milled-keys.jpg} 
  \caption{} 
  \Description{} 
  \label{fig:milling}
\end{figure}

A redesign of the PCBs meant that removing the material would no longer
be necessary.

The PCBs shared power and data lines for components with the exception
of the output from the multiplexer. Ribbon cable was cut to size to
allow for some room to readjust the distance of the sensors from the
jacks. The data signals from each PCB were connected using a separate
cable loom. Soldering cables and inspecting each join required another
day work. This did allow for readjustments to the position of the
controller board Measurements of the rear chamber did not match with the
Mac Mini that was to be used a different layout was decided for fitting
the controller board/ Cutting cables to size allowed for a great deal of
flexibility but at the cost of time. After the first install it was
clearer what constraints were in place for fitting and an alternative
wiring system using flat flex cable was implement (discussed in
\textbackslash section)

Fitting PCBs required the removal of all keys. The support brackets were
assembled and screwed into the the underside. With the harpsichord on
it's back the sensors were tested to ensure an expected range.
Self-tapping round head screws were used to to attach the brackets. The
screws were not over tightened to allow for brackets to move. The
brackets were not easily adjusted once the key were replaced. Adjusted
to a best average with subsequent adjustment carried out on the
microcontroller.

As the PCBs were fitted the jacks had the gradient stickers applied to
them. The jacks had been coated with a varnish on the short side to
provide a better surface for the stickers to adhere to. After the
stickers were applied excess material was trimmed from each jack. The
jack were then check that they moved freely in their slot.

After fitting the PCBs preliminary calibration was carried out. Before
the keys were refitted the jacks were put back in place. The data from
each key was confirmed to be in a similar range. Any key showing
anomalous data had it's jack removed and re-inspected. The PCB fitting
process and refitting required another day as faults were found in
wiring or misalignment of the stickers. After the sensors were
confirmed to be functioning the keys were put back.

The calibration workflow was with \anon{Dr. Craig Webb} who was unfamiliar with
the system. Adjustments were made to the firmware to to make it more
intuitive for the user. Addressable LEDs became vital as it allowed for
easy identification of the key being calibrated. They were also used to
identify any keys whose current reading was beyond what was expected.

The rotary encoder was used as the interface to select a key, adjust
it's threshold and update the Ferroelectric RAM (FRAM) with new values.

A process was implemented where the key was selected by playing, but
this caused problems in situations where the data was unusual due to
misalignment of the sensors.

This would likely be a better system for when the keyboard was fully
operational.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/serial_monitor.png} 
  \caption{} 
  \Description{} 
  \label{fig:serial_monitor}
\end{figure}

The thresholds were calibrated visually utilising the serial plotter of
the Arduino IDE (Figure \ref{fig:serial_monitor}) and the open source MIDI
Monitor software \footnote{\url{https://github.com/krevis/MIDIApps}}.

\begin{itemize}
\item
  partial pre-assembly
\item
  handsolder components

  \begin{itemize}
  \item
    analog lines
  \end{itemize}
\item
  connecting supports
\end{itemize}

>>>>

Technological advances have transformed how museums document, present and interpret their collections, fostering immersive experiences through tools such as laser scanning, 3D printing, and virtual reality \cite{allard2005use,Wachowiak01082009,RCM_2024_3D,Kuzminsky_LaserScan_2012,Schaich_3D_2007}. These technologies emphasise experiential authenticity, enabling encounters that evoke the past's sensory, emotional, and intellectual essence \cite{trant_Auth_1999}. However, as Pine and Gilmore note \cite{pinegilmore_2007}, achieving authenticity requires museums to navigate the delicate balance between preservation and meaningful engagement—a challenge that is particularly evident in the case of historical musical instrument collections \cite{McAlpine2014}.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{src/images/details.jpg}
\caption{Details of the replica, including the key slots and sides, keys, and jacks with seagull plectra.}\label{fig:details}
\end{figure*}

@@

The final design, visible in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key harpsichord replica equipped with an optical sensor system reading the amount of light reflected by gradient stickers applied to the jacks' sides. Several key features were implemented to address the project’s requirements:

@@

A modular system of printed circuit boards (PCBs) was designed to manage the sensors and process their output. The 49 QRE1113 sensors were distributed across seven boards, containing 7 sensors each. The PCBs were secured to the underside of the keyboard mechanism, allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

The project built upon earlier NIME research on generating MIDI messages from piano keystrokes \cite{McPherson2013}, adapting it to address the unique characteristics of harpsichords and the specified constraints. Whereas the previous design emphasised continuous gesture tracking, this implementation required discrete key-triggered data to align with the needs of MIDI-triggered audio playback.

@@

The keyboard was designed to replicate the tactile and aesthetic sensations of playing an antique Italian harpsichord. Traditional materials were used, including walnut for the wrestplank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the Alessandro Trasuntino instrument at the San Colombano Museum.

The rectangular poplar frame deviates from the traditional logarithmic form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string orders, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations. The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile and auditory qualities.

Here’s the extended **Hardware Design** section in LaTeX, revised to meet a target of approximately 700 words while maintaining a coherent and logical structure:

\section{Hardware Design}\label{hardware-design}

The hardware design aimed to fulfil the constraints identified earlier, including non-invasiveness, reliability, and scalability, while ensuring a faithful recreation of the tactile response and auditory characteristics of historical harpsichords. To achieve these goals, the system evolved through iterative prototyping, beginning with simple threshold-based testing and culminating in a fully functional multi-sensor interface capable of triggering MIDI events.

\subsection{Prototyping and Key Model}

The initial stage of development focused on testing whether sensor data could reliably trigger MIDI playback. Modifying an existing harpsichord for testing was considered but ultimately discarded due to significant discrepancies in internal measurements and layout. Instead, the team employed a custom 3-key harpsichord mechanism, crafted by Graziano Bandini and provided by San Colombano (Figure \ref{fig:3key}), as a foundation for prototyping. This approach followed a methodology similar to that used in Timmermans' Haptic Key project \cite{Timmermans2020}, enabling incremental validation of hardware and sensor integration.

The 3-key model facilitated iterative testing of individual components, including sensor placement, signal processing, and mechanical tolerances. The model’s compact form allowed for rapid experimentation while preserving the physical properties of a full-scale harpsichord.

@@

To ensure the system's suitability for a museum context, the following criteria were established to guide sensor selection and integration:

\begin{itemize}
    \item \textbf{Non-invasive:} Sensors must not require significant modifications to the harpsichord's mechanical structure or aesthetics.
    \item \textbf{Low Latency:} Sensor response times were required to remain under 10 ms, with an upper limit of 25 ms deemed acceptable \cite{Jack2016}.
    \item \textbf{Reliability:} Sensor data had to be accurate, minimising false positives and negatives without excessive filtering.
    \item \textbf{Scalability:} The design needed to be cost-effective and adaptable for up to 50 or more keys while maintaining consistent performance.
    \item \textbf{Expandability:} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

These criteria ensured that the design aligned with the open-source ethos of the project, allowing assembly in standard maker spaces without specialised equipment.

\subsection{Sensor Board Design}\label{sensor-board}

The final sensor system utilised QRE1113 infrared sensors, known for their precision and suitability for short-distance detection \cite{McPherson2013, McPherson2019}. The sensors were distributed across seven printed circuit boards (PCBs), each responsible for seven keys. Each PCB contained the following components:

\begin{itemize}
    \item 7 QRE1113 infrared sensors.
    \item 7 100 $\Omega$ resistors and 7 10 k$\Omega$ resistors (later reduced to one resistor per board).
    \item 1 Texas Instruments CD4051BE multiplexer for signal aggregation.
\end{itemize}

The QRE1113 sensors function by detecting infrared light reflected from nearby surfaces. Gradient stickers were affixed to each jack to provide a consistent reflective surface for the sensors, enabling precise tracking of jack displacement. This approach was chosen for its simplicity and scalability, allowing the design to progress from the 3-key prototype to a 49-key model without significant modifications.

To eliminate cross-talk between adjacent sensors, 3D-printed baffles were installed on the PCBs (Figure \ref{fig:baffles}). These baffles, fabricated from dark pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

The controller board, designed around the Arduino Nano BLE platform, served as the central hub for processing sensor data and triggering MIDI messages. The Nano BLE was selected for its compact form factor, integrated BLE MIDI functionality, and native USB support. These features simplified development while ensuring compatibility with existing MIDI-based synthesis systems.

Initial iterations used 2.54 mm pitch headers for cable connections, but issues with cable disconnections led to the adoption of JST-PH connectors in later designs. Custom cable looms were constructed using adapted crimping tools, ensuring robust connections without compromising accessibility.

Non-volatile memory, implemented using Ferroelectric RAM (FRAM), provided a reliable means of storing calibration data. This ensured that sensor thresholds and other settings could be preserved across power cycles, enhancing the system's usability in a museum setting.

\subsection{Calibration and Power Management}\label{calibration}

Calibration was a critical component of the hardware design. Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software (Figure \ref{fig:serial_monitor}).

Power requirements for the system were estimated at 1.1 A at 5 V, with fluctuations during startup. While the sensors were powered continuously in this iteration, future designs may incorporate power-saving measures, such as dynamic activation of sensors based on usage.

\subsection{Final Assembly and Integration}

The final 49-key harpsichord replica incorporated the sensor and controller boards into a rectangular frame, deviating slightly from traditional harpsichord designs to accommodate electronic components. Ribbon cables connected the PCBs, while felt strips were interwoven between the strings to suppress vibrations and maintain the tactile response of the keys.

The assembly process, conducted at the NEMUS lab in Bologna, involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs, as well as varnishing the jacks to improve sticker adhesion. Subsequent iterations addressed these challenges, refining the system’s scalability and robustness.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
    \caption{Cross-section illustrating sensor placement relative to the jack mechanism.}
    \label{fig:cross-section}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{3D-printed baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/serial_monitor.png}
    \caption{Arduino IDE serial plotter used for calibration of sensor thresholds.}
    \label{fig:serial_monitor}
\end{figure}

% \section{Hardware Design}

% For the first iteration the data from the sensor would simply be used to test when a threshold had been passed. On passing the threshold, a message will be sent to trigger the playback of a corresponding note.

% For the initial prototyping stage, modifying an existing harpsichord was considered though ultimately discarded. An existing instrument would
% have provided a test for scaling the electronics, but internal measurements and layout would have been too different.

% Using a similar approach to Timmermans' Haptic Key project, \cite{Timmermans2020}, our first step was to fabricate or procure a model of a single working key. San Colombano provided a model 3-key of a harpsichord mechanism by Graziano Bandini (Figure \ref{fig:3key}).

% \subsection{Sensor Criteria}\label{sensor-criteria}

% A set of criteria was drawn to which the final sensor system would have to satisfy. These criteria were used to evaluate the feasibility throughout the prototyping phase.

% \begin{itemize}
% \item
%   Non-invasive: The sensor system should not require major modifications
%   to the interface on which it is installed.
% \item
%   Low Latency: Reading sensors and then triggering an output should have
%   a low latency. Following \cite{Jack2016} an latency was set as 10 ms
%   with an upper limit set to 25 ms beyond which the sensors would be
%   deemed unusable
% \item
%   Reliable: Data obtained should be reliable and accurate enough not to
%   result in false positives or false negatives without extensive
%   filtering
% \item
%   Scalable: The system should be scalable both economically and
%   temporally. Given the open source nature of the project, it could not
%   be prohibitively expensive to carry out. Also any system that would
%   work on a single key would have to scale financially to potentially 50
%   or more. The time required to install and calibrate the sensors would
%   also have to scale.
% \item
%   Expandable: Any sensor system should be functionally expandable to
%   allow for the usage of the data for control over other parameters.
% \end{itemize}

% The sensor system was designed so that they would not require any equipment outside what can commonly be found in a maker space at a university for assembly. This decision was motivated by the project's goal to be open source and reasonably recreated without procuring additional specialist equipment.

% The electronics system can be divided into two designs, the "sensor board" (Figure \ref{fig:7-sensor-board}) for components for detecting jack displacement and controller board for collecting and processing sensor signals.

% \begin{figure}  
%   \centering
%   \includegraphics[angle=-90,width=\linewidth]{src/images/7-sensor-board-v0.1.0.png} 
%   \caption{} 
%   \Description{} 
%   \label{fig:7-sensor-board}
% \end{figure}

% \subsection{Sensor board}\label{sensor-board}

% The 49 optical sensors are divided across 7 PCBs each containing

% \begin{itemize}
% \item
%   x7 QRE1113 optical sensors
% \item
%   x7 100 $\Omega$ resistors
% \item
%   x7 10 k$\Omega$ resistors \footnote{a update on the design showed these could be reduced to a single resistor.}
% \item
%   x1 CD4051BE Multipexer
% \end{itemize}

% A number of sensors were tried before the final system was decided upon. 
% Sensors tested for the design included, a 6-degrees-of-freedom inertial measurement unit, reed and hall sensors paired with magnets, and light-dependant resistors mounted to the jack rail.
% None of the above satisfied the sensor criteria described and were thus discounted.

% Following from the work in \cite{McPherson2013, McPherson2019}, we tested
% a system using the Fairchild QRE1113 \footnote{Fairchild, now ON Semiconductor, QRE1113 datasheet \url{https://www.onsemi.com/download/data-sheet/pdf/qre1113-d.pdf}}. The
% QRE1113 is a combination infrared LED and phototransistor sensitive to
% IR light. The phototransistor of QRE1113 provides data on how much light
% is present and in particular how much light is being reflected from a
% surface. Since reflected light will be proportional to the distance of
% a surface it is a good, close-proximity, distance sensor. Distance is
% the way in which the instruments in \cite{McPherson2013, McPherson2019}
% function as well the Moog Piano Bar from which it took inspiration.

% The limitation for taking a distance approach is that there is not one
% but 2 jacks per key. For each jack to be measured independently the data
% would need to come from the jacks themselves.

% % They \emph{could} be used and individual trigger points set in
% % calibration, but this creates new problems.

% \begin{figure}  
%   \centering
%   \includegraphics[,width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:cross-section}
% \end{figure}

% The optical sensors were directed at a greyscale gradient printed on a sticker and applied to the side of the jack.
% This was first trialled using a generic inkjet printer and double-sided tape. 
% For the final iteration, the stickers were printed on Coala 1D 100 Gloss PG 100 $\mu$m white gloss monomeric PVC with permanent grey acrylic dispersion adhesive \footnote{Datasheet: \url{https://www.antalis.co.uk/mediashare/g4media/pdf/TS_EN_COALA_1D_100_GLOSS_P_00_ISS_15022023.pdf}} using a HP Latex 115 and then cut with a Summa 150 roll cutter. 
% The printer and roll cutter greatly reduced the manual labour for cutting stickers meaning the process would scale from the 6 jacks of the 3-key to the 98 jacks of the 49-key model.

% \begin{figure}  
%   \centering
%   \includegraphics[angle=90,width=\linewidth]{src/images/jack-w-tag.jpeg} 
%   \caption{} 
%   \Description{} 
%   \label{fig:jack-w-tag}
% \end{figure}

% During testing it was found that their was a considerable amount cross-talk between adjacent sensors. 
% This meant the reading from key would be affected by the movement of the key next to it. 
% The cross-talk was assumed to be as a result of light reflecting from other jacks.
% Given that IR is used this was difficult to confirm by sight. 
% A set of baffles was designed for the PCB that would wrap round the jack and stop light from other jacks or sensors \ref{fig:baffles}. 
% The baffles were printed on a BAMBU X1 3D printers using PLA plastic. A dark pigmented PLA was used to avoid IR light simply being diffused. 
% After the baffles were fitted, the cross-talk was eliminated.

% \begin{figure}  
%   \centering
%   \includegraphics[,width=\linewidth]{src/images/baffles.png} 
%   \caption{} 
%   \Description{} 
%   \label{fig:baffles}
% \end{figure}

% Calibration to pluck thresholds were to be carried out manually.  To allow for easy identification of the sensor being addressed during the calibration process, RGB LEDs were integrated into the design. The LEDs were also used in the identification of any malfunctioning or un-calibrated sensor.

% TLC5940 or other PWM driver such as in \cite{McPherson2013} add robustness should an LED fail, but complexity to assembly process. 
% Instead, the design uses addressable LEDs with integrated driver that share the same data line. 
% This has the limitation where a faulty LED will stop all subsequent LEDs in the line from working.

% The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
% We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
% An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
% The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
% Transistor / Op-amp circuit could have been used similar to the one in \cite{McPherson2013}.

% \subsection{Controller Board}\label{controller-board}

% Controller board designed around Arduino nano form factor. The Nano is
% connected through headers to allow easy switching between chips. Should
% another chip in a Nano for factor be found it could supplant the current
% Nano without any need for a redesign.

% The first version of the board exclusively used 2.54 mm pitch headers.
% The reasons for this were flexibility when creating cable looms and to
% leave open the possibility of rewiring the PCB.

% This was the case for the LEDs as they initially were to be powered on
% 5V. It was discovered that the 3V regulator of the nano was sufficient
% to power all LEDs at a lower brightness while reducing current draw.
% Since brightness was not important.

% The 2.54 mm pitch headers had two major drawbacks. First, the orientation
% of the cables could be reversed, which in the best case scenario would
% simply mean the system did not work correctly and in the worst case it
% could potentially damage the components.

% Secondly, the headers do not connect securely and wait of the cabling
% was enough for components to become disconnected as the harpsichord was
% handled.
